Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-03-04 paper/sec/conclusion.tex
minor changes to related_work and conclusion

\section{Conclusion}\label{sec:conclusion}
We presented the first formalization of the CVM algorithm using Isabelle/HOL.
Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

>>>>

\section{Conclusion}\label{sec:conclusion}
We presented the first formalization of the CVM algorithm using Isabelle/HOL.
Central to our formalization is a novel invariant-based proof technique for randomized algorithms, which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

### 2025-03-04 paper/sec/related_work.tex
minor changes to related_work and conclusion

Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented increasingly sophisticated solutions.
The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity is rather large.
What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.

@@

There is also research on the (automated) analysis of loop invariants, for probabilistic loops, using their characteristic functions~\cite{batz2023, mciver2005}.
This approach works by establishing the limiting distribution of the state of the loop.
De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
Our approach differs from these techniques, by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
Instead, we investigate invariants of classes of functions of the distributions, which are relevant for the analysis.
There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
However, these methods, do not extend to algorithms with branches or, more generally, algorithms which contain discrete operations.

Finally, verification of randomized algorithms has been tackled by authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
The most closely related efforts are the mechanizations of frequency moments algorithms by Karayel~\cite{karayel2022, karayel2023}.
The functional invariant proof technique we introduce here should be applicable in any higher-order setting.

>>>>

Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented increasingly sophisticated solutions.
The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.

@@

There is also research on the (automated) analysis of loop invariants, for probabilistic loops, using their characteristic functions~\cite{batz2023, mciver2005}.
This approach works by establishing the limiting distribution of the state of the loop.
De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
Our approach differs from these techniques, by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
Instead, we investigate invariants of classes of functions of the distributions, which are relevant for the analysis.
There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
However, these methods do not extend to algorithms with branches or, more generally, algorithms which contain discrete operations.

Finally, verification of randomized algorithms has been tackled by various authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
The most closely related efforts are the mechanizations of frequency moments algorithms by Karayel~\cite{karayel2022, karayel2023}.
The functional invariant proof technique we introduce here should be applicable in any higher-order setting.
